---
title: "Project Report"
author: "Ye Zhang"
date: "9/11/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Outline

This project is to perform an analysis of the statistical analyses in all published [PLoS](https://www.plos.org) papers, so as to answer quenstions as below:

* What are the most common techniques? 
* How do they vary by field? 
* Are there any trends over the last 10-15 years?

**Project Step 1**

Data: the dataset for this project should include all the published [PLoS](https://www.plos.org) papers from its 7 journals, [PLoS one](http://journals.plos.org/plosone/), [PLoS Biology](http://journals.plos.org/plosbiology/), [PLoS Medicine](http://journals.plos.org/plosmedicine/), [PLoS Comutational Biology](http://journals.plos.org/ploscompbiol/), [PLoS Genetic](http://journals.plos.org/plosgenetics/), [PLoS Neglected Tropical Diseases](http://journals.plos.org/plosntds/) and [PLos Pathogens](http://journals.plos.org/plospathogens/). For each publication, there'are a list of information we need to download from the websites into our R program as the dataset:

* Text of the article (including figure legends)
* Article title
* Authors
* PLoS journal
* Date of publication

**Project Step 2**

Usually the statistical analysis techique utilized in a publication is described in the *Materials and Methods* section of the article,thus we should focus on 



**Project Step 3**



```{r }

```

